# Monitoring Early Onset Crises with NTL

At the onset of war in Ukraine it was difficult to obtain accurate information regarding the conditions on the ground. Therefore, a study of Night Time Light luminosity was conducted. The analyses revealed significant shifts in night-time lights over the coarse of the invasion. These changes can be attributed to:

- damage to infrastructure
- change in human activity connected to norms established through martial law
- change in human activity connected to the ongoing evacuations and relocation of residents to other parts of the country or neighboring cities

Night time lights data can be obtained from several sources, but the blackmarble VNP46A suite images was required for this work to obtain and deliver analyses in near real time (NRT)

## Downloading

The [NASA - LAADS WorldView Portal](https://worldview.earthdata.nasa.gov/) was used to query and download raster images. The Black Marble Nighttime Blue/Yellow Composite (DNB) which combines the Black Marble TOA nighttime radiance  product with infrared bands was essential for querying and filtering the images to detect influence of moonlight and clouds. An excellent [article](https://www.earthdata.nasa.gov/learn/articles/black-marble-blue-yellow-composite) by Joseph M. Smith was used as the basis of this technique. 


- Use blue-yellow composite imagery to better understand cloud formations & atmospheric effects
- If downloading data from the last ~ 72 hours you will need to DL from the NRT collection
- It is possible that you will get a message saying the imagery from the NRT collection is not available. If this happens it is because you are somehow logged into LAADs system, but not the NRT catalog. In this case go to the [NRT Portal](https://nrt3.modaps.eosdis.nasa.gov/archive/allData/5000/VNP46A1G_NRT/2022/073) and click log in.



To facilitate the processing and reprojecting of `.h5` files obtained from the world view portal the `{blackmarble}` R package was developed based on [script provided by NASA](https://git.earthdata.nasa.gov/projects/LPDUR/repos/nasa-viirs/browse/scripts)

The development version of blackmaRble can be downloaded from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("zackarno/blackmaRble")
```

After filtering images save all the `.h5` format raster tiles to 1 directory with the default NASA file name syntax.

## Processing


```{r eval=F}
library(tidyverse)
# devtools::install_github("zackarno/blackmaRble")
library(blackmaRble)
library(here)
library(terra)
write_rasters <-  c(T,F)[1] # toggle 1 for write outputs, 2 for no outputs
```

### Reprojecting

- using list of file names & list of of full path names we can read in all the files in the folder and reproject them all at once
- results are saved in list object: `ntl_reprojected`

```{r pressure, echo=T, eval=F}
# dir.create("ntl_h5")
h5_filenames <- list.files(here::here("ntl_h5/"),pattern = "\\.h5$")
h5_paths <- list.files(here::here("ntl_h5/"), full.names = T,pattern = "\\.h5$")
ntl_reprojected <- h5_paths |> 
  map(~vnp46A_to_wgs84(path =.x)) |> 
  set_names(h5_filenames)
```

### Subset bands

- to simplify list object lets just subset each item in list to the DNB band
- new object `dnb_radiance` contains only `DNB_At_Sensor_Radiance_500m` band for each item in list.

```{r , eval=F}
# subset to just main band of interest for faster mosaicing
dnb_radiance <- ntl_reprojected |> 
  map(~.x |> subset("DNB_At_Sensor_Radiance_500m"))
```


### Mosaic & write rasters 
- since study area is composed of 4 tiles we need to mosaic them together
- function below uses NASA naming syntax to group tiles together by `DOY` and then mosaic them
- specify `output` to automatically write as tiff to specified directory
- result is also saved as `spatRaster collection`

```{r,eval=F}
# dir.create("ntl_doy_tiffs")
if(write_rasters){
  ntl_rad_by_doy <-  mosaic_vnp46a_by_doy(dnb_radiance,output = here::here("ntl_doy_tiffs"))
}
# ntl_rad_by_doy$`084`
```


## Use-Case Examples

[Impact-Initiatives Ukraine Night-time Lights Pre and Post Conflict Escalation - February vs March 2022](https://www.impact-repository.org/document/impact/9718da8e/impact_ukraine_map_nighttime-lights_17Mar2022_a1.pdf)
